Detection and replication of epistasis influencing transcription in humans

### 2013-09-16 docs/manuscript/manuscript.tex
changed some things in response to peter's comments

\author[1,2,*]{Gibran Hemani}
\author[1,2]{Konstantin Shakhbazov}
\author[3]{Harm-Jan Westra}
\author[4,5,6]{Tonu Esko}
\author[7]{Anjali K Henders}
\author[1,2]{Allan F. McRae}
\author[2]{Jian Yang}
\author[8]{Greg Gibson}
\author[7]{Nicholas G Martin}
\author[4]{Andres Metspalu}
\author[3]{Lude Franke}
\author[7,+]{Grant W Montgomery}
\author[1,2,+]{Peter M Visscher}
\author[1,2,+]{Joseph E Powell}

@@

Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome. The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and examples have been reported in other species \cite{Carlborg2006}, few convincing examples with independent replication exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs). In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$). Replication of these signals in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 501$. There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs. For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first evidence for multiple instances of natural genetic polymorphisms interacting to influence human traits.

@@

Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power. For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects. Thus, when combined with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two simple ways to overcome this problem. One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes among common variants. Because our focus was to ascertain the extent to which instances of epistasis occur amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes. These traits are largely heritable but on average have been shown to be less polygenic than high level phenotypes \cite{Powell2013}, thus many genetic effects are relatively large \cite{Powell2012}, maximising the chance at detecting epistasis, should it exist.

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions. First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood (family-wise error rate of 5\% corresponding to a significance threshold of $ p < 2.91 \times 10^{-16}$, Methods). Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $\alpha \approx 0.07$, Methods, Supplementary Figure \ref{fig:type1}). Using this design we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}). Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/501$, Table \ref{tab:bonferroni}). To quantify the similarity of GP maps between the independent datasets (Figure \ref{fig:gpmaps}) we decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A \times A$, $A \times D$, $D \times A$, $D \times D$) and tested for concordance of the sign of the most signicant effect (Supplementary Table \ref{tab:sign}, Methods). Sign concordance between the discovery and both replication datasets was observed in 22 out of the 30 significantly replicated interactions (expected value $= 7.5$ under the null hypothesis of no interactions, $p = 3.76 \times 10^{-8}$).

@@

It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is difficult in practice for several reasons. For example, LD between causal variants and observed markers plays an important role. Not only is the dependence on LD much greater for epistatic effects than for additive effects (Supplementary Figure \ref{fig:ld_sampling_mean}), but when estimating epistatic variance it is more sensitive to changes in LD between observed SNPs and causal variants between independent samples when compared to additive effects (Supplementary Figure \ref{fig:ld_sampling_sd}), and this has a direct effect on statistical power for replication. Indeed, in addition to heterogeneity between populations, sampling variance of LD $r$ alone leads to over three fold decrease in observed variance for the most extreme case of $D \times D$ effects when compared to the average decrease for additive variance on average (Methods, Supplementary Figure \ref{fig:ld_reduction}). Therefore these results are encouraging with regards to the detection and replication of epistasis.

@@

% \subsection{Contribution of epistatic variance}
Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question. In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance. Taking published results from a standard eQTL study \cite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the additive variance of a locus was $2.1\%$ or greater. Thus, we can infer that the number of instances of large additive effects is significantly greater than the number of instances of large epistatic effects.

In terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}. Ideally one would approach this question from a whole genome perspective \cite{Visscher2008} but this is intractable for non-additive variance components. Nevertheless, some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}. Taking the additive effects detected in Powell \emph{et al} (2012) where the minimum additive variance was $2.1\%$, we calculated that the proportion of total phenotypic variance of all 7339 gene expression levels explained by additive effects alone was $2.16\%$. By contrast, the epistatic variance from the interacting SNPs detected in this study on average explain a total of 0.22\% of phenotypic variance, approximately ten times lower than the additive variance. There are several caveats to this comparison. Firstly, the ratio of additive to epistatic variance may differ at different minimum variance thresholds, and our estimate is determined by the threshold used. Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test. And thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance. This is due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.

@@

\subsection*{Methods Summary}
We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs. Each individual had gene expression levels measured in peripheral blood at 47,323 probes. Only the probes that passed quality control and had significant expression in $\geq 90\%$ of individuals were used in the analysis (7,339 probes representing 6,158 RefSeq genes). Recent hardware and software \cite{Hemani2011} advances that use graphics processing units (GPUs) made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis. We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $T_{e} = 2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER). SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms. Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}. Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone. All SNP pairs with LD $r^2 > 0.1$ and $D'^{2} > 0.1$ were removed to minimise the possibility of haplotype effects. All SNP pairs were required to have at least five data points in all nine genotype classes. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained. Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs. The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$). Of these, 434 passed filtering in both replication datasets. A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{type1}
	\caption{\textbf{Type 1 error rate of two stage design} In stage 1 SNPs are tested for full genetic effects (8 d.f.) and those that surpass a threshold for multiple testing are then tested for significant interaction terms in stage 2. These interaction $p$-values are then adjusted (Bonferroni) for the total number of tests that passed stage 1. The type 1 error rate of this two stage design is dependent on the power, which is not known empirically.}
\label{fig:type1}
\end{figure}
\clearpage

@@

\begin{figure}
	\centering
	\includegraphics[width=5in]{ld_sampling_mean}
	\caption{\textbf{Sampling mean for different power terms of population $r$ values} Power of detection and replication of epistatic signals depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$. For a given a population value of LD $r$ (columns of plots), plotted is the sample mean ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines). As true $r$ reduces the statistical power to detect epistatic variants drops dramatically under the assumption that statistical power is proportional to higher moments of $r$.}
	\label{fig:ld_sampling_mean}
\end{figure}
\clearpage

\begin{figure}
	\centering
	\includegraphics[width=5in]{ld_sampling_sd}
	\caption{\textbf{Sampling standard deviation for different power terms of population $r$ values} Power of detection and replication of epistatic signals depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$. For a given a population value of LD $r$ (columns of plots), plotted is the sampling standard deviation ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines). As the power term of $r$ increases the sampling variance also increases. Supposing that there is sufficiently high $r^x$ in the discovery sample for detection of epistasis, the replication sample is less likely to have similarly high $r^x$ as $x$ increases, leading to an expectation of reduced replication rates.}
	\label{fig:ld_sampling_sd}
\end{figure}
\clearpage

>>>>

\author[1,2,*]{Gibran Hemani}
\author[1,2]{Konstantin Shakhbazov}
\author[3]{Harm-Jan Westra}
\author[4,5,6]{Tonu Esko}
\author[7]{Anjali K Henders}
\author[1,2]{Allan F McRae}
\author[2]{Jian Yang}
\author[8]{Greg Gibson}
\author[7]{Nicholas G Martin}
\author[4]{Andres Metspalu}
\author[3]{Lude Franke}
\author[7,+]{Grant W Montgomery}
\author[1,2,+]{Peter M Visscher}
\author[1,2,+]{Joseph E Powell}

@@

Epistasis is the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome. The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and examples have been reported in other species \cite{Carlborg2006}, few convincing examples with independent replication exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs). In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$). Replication of these interactions in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 501$. There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs. For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first evidence for multiple instances of segregating common polymorphisms interacting to influence human traits.

@@

Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power. For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects. Thus, when combined with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two simple ways to overcome this problem. One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes among common variants. Because our focus was to ascertain the extent to which instances of epistasis arises from natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes and like most complex diseases, these expression traits are typically heritable \cite{Schadt2003}. But unlike complex diseases, genetic associations with gene expression commonly have very large effect sizes that explain large proportions of the genetic variance\cite{Powell2013}, making them good candidates to search for epistasis, should it exist.

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions. First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood (family-wise error rate of 5\% corresponding to a significance threshold of $ p < 2.91 \times 10^{-16}$, Methods). Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $0.05 \leq \alpha \leq 0.14$, Methods, Supplementary Figure \ref{fig:type1}). Using this design we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}). Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/501$, Table \ref{tab:bonferroni}). To quantify the similarity of GP maps between the independent datasets (Figure \ref{fig:gpmaps}) we decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A \times A$, $A \times D$, $D \times A$, $D \times D$) and tested for concordance of the sign of the most signicant effect (Supplementary Table \ref{tab:sign}, Methods). Sign concordance between the discovery and both replication datasets was observed in 22 out of the 30 significantly replicated interactions (expected value $= 7.5$ under the null hypothesis of no interactions, $p = 3.76 \times 10^{-8}$).

@@

It should be noted that although it is a necessary step to establish the veracity of the interactions from the discovery set, replication of epistasis is difficult in practice. For example, LD between causal variants and observed markers plays an important role. Not only is the dependence on LD much greater for epistatic effects than for additive effects (Supplementary Figure \ref{fig:ld_sampling_mean}), but when estimating epistatic variance it is more sensitive to changes in LD between observed SNPs and causal variants between independent samples when compared to additive effects (Supplementary Figure \ref{fig:ld_sampling_sd}), and this has a direct effect on statistical power for replication. The sampling variance of LD $r$ leads to the ascertainment of marker associations with higher sample $r$ in the discovery stage in comparison to the replication stage. However, the average decrease in $\hat{r}^{x}$ in replication samples becomes larger as $x$ increases (Methods, Supplementary Figure \ref{fig:ld_reduction}). For example, the decrease in $\hat{r}^8$ (which is proportional to the power of deteting $D \times D$ effects), is on average three fold greater than the decrease in $\hat{r}^2$ (which is proportional to the power of detecting additive effects).

@@

% \subsection{Contribution of epistatic variance}
Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question. In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum estimated variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance. Taking results from our previously published eQTL \cite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the estimated additive variance of a locus was $2.1\%$ or greater. Thus, we can infer that the number of instances of large additive effects is significantly greater than the number of instances of large epistatic effects.

In terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}. Ideally one would approach this question from a whole genome perspective \cite{Visscher2008} but this is intractable for non-additive variance components. Nevertheless, some inference can be made from the ascertained effects in these analyses and it is evident that estimated additive variance is overall a larger component than estimated epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}. Taking all additive effects detected in Powell \emph{et al} (2012) that have additive variance explaining $2.1\%$ or greater of phenotypic variance, we calculated that the proportion of total phenotypic variance of all 7339 gene expression levels explained by additive effects alone was $2.16\%$. By contrast, the estimated epistatic variance from the interacting SNPs detected in this study on average explain a total of $0.22\%$ of phenotypic variance, approximately ten times lower than the estimated additive variance. There are several caveats to this comparison. Firstly, the ratio of additive to epistatic variance may differ at different minimum variance thresholds, and our estimate is determined by the threshold used. Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test. Thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance. This is due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs. And forthly, the extent of winner's curse in estimation of effect sizes may differ between the the two studies.

@@

\subsection*{Methods Summary}
We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs. Each individual had gene expression levels measured in peripheral blood at 47,323 probes. Only the probes that passed quality control and had significant expression in $\geq 90\%$ of individuals were used in the analysis (7,339 probes representing 6,158 RefSeq genes). Recent hardware and software \cite{Hemani2011} advances that use graphics processing units (GPUs) made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis. We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $T_{e} = 2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER). SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms. Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}. Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone. All SNP pairs with LD $r^2 > 0.1$ and $D'^{2} > 0.1$ were removed to minimise the possibility of haplotype effects. All SNP pairs were required to have at least five data points in all nine genotype classes. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinel SNP pair was retained. Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs. The 501 significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in peripheral blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$). Of these, 434 passed filtering in both replication datasets. A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{type1}
	\caption{\textbf{Type 1 error rate of two stage design assuming a null model of one large additive effect and no epistasis} In stage 1 SNPs are tested for full genetic effects (8 d.f.) and those that surpass a threshold for multiple testing are then tested for significant interaction terms in stage 2. These interaction $p$-values are then adjusted (Bonferroni) for the total number of tests that passed stage 1. The type 1 error rate of this two stage design is dependent on the power, which is not known empirically.}
\label{fig:type1}
\end{figure}
\clearpage

@@

\begin{figure}
	\centering
	\includegraphics[width=5in]{ld_sampling_mean}
	\caption{\textbf{Sampling mean for different power terms of population $r$ values} Power of detection and replication of epistatic interactions depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$. For a given population value of LD $r$ (columns of plots), plotted is the sample mean ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines). As true $r$ reduces the statistical power to detect epistatic variants drops dramatically under the assumption that statistical power is proportional to higher moments of $r$.}
	\label{fig:ld_sampling_mean}
\end{figure}
\clearpage

\begin{figure}
	\centering
	\includegraphics[width=5in]{ld_sampling_sd}
	\caption{\textbf{Sampling standard deviation for different power terms of population $r$ values} Power of detection and replication of epistatic interactions depends not on $r^2$ between causal variants and observed SNPs, but on $r^4, r^6, r^8$. For a given a population value of LD $r$ (columns of plots), plotted is the sampling standard deviation ($y$-axis) of $\hat{r}$, $\hat{r}^2$ (additive), $\hat{r}^4$ (dominance, A$\times$A), $\hat{r}^6$ (A$\times$D), $\hat{r}^8$ (D$\times$D) ($x$-axis) for different sample sizes (coloured lines). As the power term of $r$ increases the sampling variance also increases. Supposing that there is sufficiently high $r^x$ in the discovery sample for detection of epistasis, the replication sample is less likely to have similarly high $r^x$ as $x$ increases, leading to an expectation of reduced replication rates.}
	\label{fig:ld_sampling_sd}
\end{figure}
\clearpage
